Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-07 paper/paper.tex
fixed a few minor typos

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}B).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

@@

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}C, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\FIGSUPP[transcriptomics]{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.

@@

It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if an infected triggers an IFN response.
We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

@@

\section{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, and Katherine Xue for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.
The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.

>>>>

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

@@

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \FIGSUPP[transcriptomics]{ISG}).
Over 20\% of  infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency of $\sim$0.5\% (\FIG{IFNrare}C) among infected cells.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\FIGSUPP[transcriptomics]{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\FIGSUPP[transcriptomics]{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not themselves detect infection~\citep{stetson2006type,honda2006type}.

@@

It turns out that these viral defects substantially contribute to the heterogeneity observed in prior single-cell studies of influenza~\citep{russell2018extreme, steuerman2018dissection, heldt2015single, sjaastad2018distinct}, including helping determine if a cell triggers an IFN response.
We identified four types of viral defects in single IFN+ cells that we validated to increase IFN induction. 
Two types of defects---absence of the NS gene and amino-acid mutation to the NS1 protein---presumably impair NS1's well-known ability to antagonize innate-immune pathways~\citep{garcia1998influenza, hale2008multifunctional}.
The third type of defect, amino-acid mutations in the PB1 protein, is consistent with recent work by \citet{velthuis2018mini} showing that mutations to the viral polymerase affect generation of mini-viral RNAs that activate the innate-immune sensor RIG-I.
Finally, we found an internal deletion in the PB1 gene that enhances IFN induction, consistent with a large body of prior work showing that such deletions are immunostimulatory~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.
Given the extensive prior work describing such deletions, it is perhaps surprising we did not identify more of them in our single IFN+ cells.
There may be several reasons: we used relatively pure viral stocks~\citep{xue2016propagation} at modest MOI; our experiments preferentially captured cells with higher viral transcriptional load; and most prior studies have used techniques that can detect large deletions but not subtle point mutations.
The relative importance of different types of viral defects for IFN induction likely varies across infection conditions, viral strains, and cell types---so it remains an open question as to which defects are most relevant for early immune detection of actual viral infections in humans.

@@

\section{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation.
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.
The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.
